Quantum H theorem, variational entropy, and relaxation processes

### 2021-03-10 tex/disscussion.tex
Removed comments (pdf remains the same)
Cambridge corrections applied.

%--------------------------------------
% Me parece que sera necesario incluir la discusin sobre los procesos de relajacin.
%Sin esa parte el contenido original del artculo podria parecer insuficiente para
%una buena publicacion%

@@

The demonstration of the classical $H$-theorem usually starts by assuming that the gas,
despite being initially out of equilibrium, can be described by a
spatially homogeneous distribution function. 
Subsequently, the time-evolution of the system occurs in such a manner that $dH/dt \leq 0$. 
Therefore, this approach does not cover the evolution to equilibrium of systems with spatial inhomogeneities.
To address this issue, in this article we
proposed a framework that may be useful to describe the time-evolution of systems
initially non-homogeneous. To this end, we divided the
system into small cells, so as to conceive a system wherein the local equilibrium
hypothesis is valid in each cell, but in such a manner that the total system is not homogeneous. Systems
that satisfy the previous conditions will evolve towards equilibrium, and the evolution
occurs according to $d\mathcal{H}'/dt\leq0$, Eq.~(\ref{eq:cHdef}),
and $d\mathcal{H}/dt\leq0$, Eq.~(\ref{eq:qHdef}), for classical and quantum gases, respectively.
Consequently, this approach can be considered as an extension of the $H$-theorem
for more realistic out-of-equilibrium systems.

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions both in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. The relaxation process of the system is described by
monotonic functions that account for deviations from the global equilibrium.

It is clear that for describing the relaxation process of a concrete system,
it is necessary to know, at least with some approximation level, the specific forms of the
monotonic functions $g'_M$ and $g_{nM}$, for classical and quantum systems.
Whereas the complete analysis of these functions is out of the scope of the present work,
some of their properties can be foreseen, \textit{e.g.} they must
be consistent both with the system relaxation times and with the mechanisms of 
energy transfer between cells.

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equlibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of dilute gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of out-of-equilibrium systems, other than dilute gases. In particular, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the explicit functional form of $g'_M$ and $g_{nM}$.

In summary, we proposed a variational procedure to demonstrate the classical and
quantum $H$-theorems, which made possible to describe, at a mesoscopic local view
(cell-scale), the time-evolution of an out-of-equilibrium and spatially
non-homogeneous system, towards the equilibrium condition. In principle, this
approach would render possible to investigate the transport phenomena, inherent to the
equilibration process, occurring in a system that started from an spatially inhomogeneous
out-of-equilibrium initial condition.

%-----------------------------------------------

>>>>

@@

The demonstration of the classical $H$-theorem usually begins by assuming that the gas,
despite being initially out of equilibrium, can be described by a
spatially homogeneous distribution function. 
Subsequently, the time-evolution of the system occurs in such a manner that $dH/dt \leq 0$. 
Therefore, this approach does not describe the evolution to equilibrium of systems with spatial inhomogeneities.
To address this issue, in this article, we
proposed a framework that may be useful to describe the time-evolution of 
initially non-homogeneous systems. To this end, we divided the
system into small cells to conceive a system wherein the local equilibrium
hypothesis is valid in each cell but in such a manner that the total system is not homogeneous. Systems
that satisfy the previous conditions will evolve towards equilibrium, and the evolution
occurs according to $d\mathcal{H}'/dt\leq0$, Eq.~(\ref{eq:cHdef})
and $d\mathcal{H}/dt\leq0$, Eq.~(\ref{eq:qHdef}), for classical and quantum gases, respectively.
Consequently, this approach can be considered an extension of the $H$-theorem
for more realistic out-of-equilibrium systems.

The classical and quantum $H$-functionals, $\mathcal{H}'$ and $\mathcal{H}$, respectively,
correctly recover the most-probable distribution
functions in out-of-equilibrium states (locally) and when the system attains the global
equilibrium state. The relaxation process of the system is described by
monotonic functions that account for deviations from the global equilibrium.

It is clear that for describing the relaxation process of a concrete system,
it is necessary to know, at least to some approximation, the specific forms of the
monotonic functions $g'_M$ and $g_{nM}$, for classical and quantum systems.
Whereas the complete analysis of these functions is beyond the scope of the present work,
some of their properties can be predicted, \textit{e.g.} they must
be consistent both with the system relaxation times and the mechanisms of 
energy transfer between cells.

An important aspect of the framework proposed in this work is related to the entropy of systems
out-of-equilibrium. Because the functionals $\mathcal{H}'$ and $\mathcal{H}$ can be related to
the entropy of dilute gases, either classical or quantum, the fact that these functionals are defined
over a system divided into cells enables their use for defining the entropy
of out-of-equilibrium systems, other than dilute gases. Specifically, and 
derived from our previous work (\textit{e.g.} \cite{bib:nicolas2020,bib:nicolas2016}), the
$\mathcal H'$ and $\mathcal H$ functionals may serve to describe the entropy,
as well as the entropy generation, occurring during the
growth of complex physical systems, such as fractals. Possibly, studying these systems might also
shed light on the explicit functional form of $g'_M$ and $g_{nM}$.

In summary, we proposed a variational procedure to demonstrate the classical and
quantum $H$-theorems, which allowed us to describe, at a mesoscopic local view
(cell-scale), the time-evolution of an out-of-equilibrium and spatially
non-homogeneous system moving towards the equilibrium condition. In principle, this
approach would permit the investigation of the transport phenomena inherent to the
equilibration process, occurring in a system with a spatially inhomogeneous
out-of-equilibrium initial condition.

### 2021-03-10 tex/introduction.tex
hice un ligero cambio en las citas
Cambridge corrections applied.

The theoretical basis and the procedures that allow to describe equilibrium systems
are well-stablished. These procedures can be applied to treat a wide range of the systems
present in nature, and include both the macroscopic phenomenological methods (thermodynamics)
as well as the microscopic description (statistical mechanics).
(Of course, regarding out-of equilibrium systems, the present situation is still a challenge.)
For instance,
in the Kinetic Theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards the equilibrium is finely accounted for
through the Boltzmann $H$-theorem.

On the other hand, for quantum out-of-equilibrium systems, the construction of a
kinetic framework, with the same level of
success and universality as the classical version, still presents some fundamental challenges. For instance,
in order to obtain a complete correspondence principle, between classical mechanics
and quantum mechanics, the quantum analogous of both the Boltzmann $H$-theorem
and the Boltzmann transport equation have not attained a fully accepted form.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using probability transition relations, the random
phases hypothesis, and an $H$-functional defined in terms of a spatially homogeneous distribution function.
Also, Tolman proposed a potential quantum analogous of the transport equation,
in terms of the occupation numbers,
by applying time perturbation theory. Further attempts, under the quantum operator
formalism, have addressed the description of quantum transport phenomena by means of the Hamiltonian
of the system and the master equation (which is, in these works, the analogous
of the Boltzmann transport equation)
\cite{bib:grabert1974,bib:wang2014,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
In the same line of thought, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed, or
if their framework fulfills the correspondence principle is unclear or not discussed whatsoever.
After the pioneering work of Tolman, at several stages, there has been some discussion regarding
the general validity of the quantum $H$-theorem,
some possible violations of the second law of thermodynamics, and the
interpretation of the quantum entropy  %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

On the other hand, the framework to describe spatially non-homogeneous systems is still
under construction, although several approaches have been developed. For instance,
the celebrated Onsager formulation (linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931} have been successful in describing irreversible chemical 
and physical phenomena. Nonetheless, some topics, such as describing the internal behavior of glasses
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018}, cannot be completely addressed with linear thermodynamics.

In addition, some aspects regarding the classical $H$-theorem and the
Boltzmann $H$-functional require to be revised, for the purpose of improving their mutual consistency.
For example, the modification of the $H$-theorem to include phenomena stemming from stochastic
trajectories, violations of the second law of thermodynamics, the relation
between Shannon's measure of information and the Boltzmann's entropy, 
and the calculation of thermodynamical quantities and thermalization of certain systems
\cite{bib:gorban2014,bib:li2019,bib:gring2012,bib:nemilov2018,bib:wang2014}.

In order to contribute to the construction of a consistent classical and
quantum $H$-theorem, within a formalism that describes out-of-equilibrium
non-homogeneous systems, in this article we propose a new theoretical framework.
Specifically, for both classical and quantum systems, we include non-homogeneous distribution functions in
the $H$-functionals, and consider non-homogeneous systems in the proofs of the resulting
$H$-theorems. Our proposed $H$-functionals satisfy
the correspondence principle, but more importantly, these functionals describe
the time-evolution of spatially non-homogeneous
systems towards equilibrium.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, to our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, followed by our proposal of
an alternative $H$-functional for classical systems, and the demonstration of
the respective $H$-theorem . In Section~\ref{sec:quantum}
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework, subsequently we present our proposal for a quantum $H$-functional,
and the proof of the corresponding quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
the classical-quantum correspondence between the
quantum and classical $H$-functionals is analyzed.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas and, on the basis of what we call variational entropy,
we propose a time-evolution equation for the distribution function. Finally,
we discuss some key ideas derived from our approach and close with a brief summary
in Section~\ref{sec:disscussion}.

>>>>

The theoretical bases and the procedures that allow us to describe equilibrium systems
are well-established. These procedures can be applied to a wide range of natural systems,
including both the macroscopic phenomenological methods (thermodynamics)
and the microscopic description (statistical mechanics).
(Out-of-equilibrium systems, of course, are still a challenge.)
For instance,
in the kinetic theory of gases, the behavior of a dilute classical gas
is described through the Boltzmann transport equation \cite{bib:huang},
and the time-evolution of a system towards equilibrium is finely accounted for
through the Boltzmann $H$-theorem.

However, for quantum out-of-equilibrium systems, the construction of a
kinetic framework with the same level of
success and universality as the classical version still presents some fundamental challenges. For instance,
to obtain a complete correspondence principle between classical mechanics
and quantum mechanics, the form of the quantum analogues of both the Boltzmann $H$-theorem
and the Boltzmann transport equation is inadequate.
In this context, Tolman was one of the earliest physicists to propose a quantum
$H$-theorem \cite{bib:tolman}, using a probability transition relationship, the random
phases hypothesis, and an $H$-functional defined in terms of a spatially homogeneous distribution function.
Tolman also proposed a potential quantum analogue of the transport equation,
in terms of the occupation numbers,
by applying time perturbation theory. Additional attempts, under quantum operator
formalism, have addressed the description of quantum transport phenomena through the Hamiltonian
of the system and the master equation (which is, in these works, the analogue
of the Boltzmann transport equation)
\cite{bib:grabert1974,%
bib:angel2017,
bib:amato2020,bib:nicacio2015,bib:hussein2014}. However, these approaches 
are not consistent with the classical-quantum correspondence principle.
Similarly, some authors have proposed $H$-functionals
and attempted to proof a
quantum $H$-theorem \cite{bib:gorban2014,bib:bennaim2017,bib:silva2010,%
bib:deroeck2006,bib:acharya2019,bib:kastner2017,bib:han2015,bib:das2018,bib:vonneumann2010}.
However, whether or not the homogeneous distribution function hypothesis is assumed or
if its framework fulfills the correspondence principle is unclear or not discussed.
Since the pioneering work of Tolman, at several stages, there has been some discussion regarding
the general validity of the quantum $H$-theorem,
some possible violations of the second law of thermodynamics, and the
interpretation of the quantum entropy  %[19-26].
\cite{bib:silva2010,bib:acharya2019,bib:kastner2017,bib:han2015,bib:brown2008,%
bib:dragoljub2009,bib:vonneumann2010,%
bib:syros1999,bib:lesovik2016,
bib:lesovik2019}.

Nonetheless, the framework to describe spatially non-homogeneous systems is still
under construction, although several approaches have been developed. For instance,
the celebrated Onsager formulation (linear thermodynamics)
\cite{bib:keizer1987,bib:onsager1931} has been successful in describing irreversible chemical 
and physical phenomena. However, some descriptions, such as those the internal behavior of gases
\cite{bib:zanotto2018} and the entropy measurement
\cite{bib:schmelzer2018,bib:nemilov2018}, cannot be completely addressed with linear thermodynamics.

In addition, some aspects regarding the classical $H$-theorem and the
Boltzmann $H$-functional require revision to improve their mutual consistency.
One example is the modification of the $H$-theorem to include phenomena stemming from stochastic
trajectories, violations of the second law of thermodynamics, the relationship
between Shannon’s measure of information and the Boltzmann’s entropy, 
and the calculation of thermodynamical quantities and thermalization of specific systems
\cite{bib:gorban2014,bib:li2019,bib:gring2012,bib:nemilov2018,bib:wang2014}.

To contribute to the construction of a consistent classical and
quantum $H$-theorem, within a formalism that describes out-of-equilibrium
non-homogeneous systems, we propose a new theoretical framework.
Specifically, for both classical and quantum systems, we include non-homogeneous distribution functions in
the $H$-functionals, and consider non-homogeneous systems in the proofs of the resulting
$H$-theorems. Our proposed $H$-functionals satisfy
the correspondence principle, but more importantly, these functionals describe
the time-evolution of spatially non-homogeneous
systems towards equilibrium.

The organization of this article is as follows.
In Section~\ref{sec:classical}, we highlight, for our purposes, the most
fundamental assumptions required to proof the Boltzmann $H$-theorem,
we provide an alternative method to obtain the Maxwell-Boltzmann
distribution using the variational method, propose
an alternative $H$-functional for classical systems, and demonstrate
the respective $H$-theorem. In Section~\ref{sec:quantum},
we review the Tolman proposal for the quantum
version of the $H$-theorem (quantum $H$-theorem)
and how the Bose-Einstein and Fermi-Dirac distributions are treated within
this framework. Subsequently, we present our proposal for a quantum $H$-functional
and the proof of the corresponding quantum $H$-theorem. In 
Section~\ref{sec:qccorrespondence}
we analyze the classical-quantum correspondence between the
quantum and classical $H$-functionals.
In Section~\ref{sec:relaxproc} we explore how relaxation processes occur
in a quantum ideal gas and, based on what we call variational entropy,
propose a time-evolution equation for the distribution function. Finally,
we discuss some key ideas resulting from our approach and close with a summary
in Section~\ref{sec:disscussion}.

### 2021-03-10 tex/qccorrespondence.tex
Cambridge corrections applied.

\section{Quantum Classical correspondence}\label{sec:qccorrespondence}

In sections \ref{sec:classical} and \ref{sec:quantum}, we have seen that the variational
method can be applied to $H$-functionals, which correctly describes the behavior of
classical and quantum dilute gases, as to their respective time-evolution.
Both $H$-functionals defined in Eqs.~(\ref{eq:cHdef}) and
(\ref{eq:qHdef}) also recovers the well-known distribution functions, either Maxwell-Boltzmann
for a classical gas, Fermi-Dirac for a Fermi gas, or Bose-Einstein for a Bose gas. Nevertheless,
the functionals (\ref{eq:cHdef}) and
(\ref{eq:qHdef}) are seemingly different, and in this section we will show that both are
related by the correspondence principle.

We start by arguing that, in equilibrium, it is straightforward to proof that
Eq.~(\ref{eq:qfneqtwo}) collapses
into Eq.~(\ref{eq:fbardef}) by taking the limit wherein the 
degeneration parameter $\xi\equiv\exp\big(-(\epsilon-\mu)/(k_BT)\big) \gg 1$.
Alternatively, a more general approach to show the quantum classical correspondence
consists of analyzing the collapse from the quantum to the classical $H$-functionals,
within the appropiated limit.
For the case treated here, this limit is $f_{nM} \approx 0$ because of the following reasons.
Systems at very low temperatures, where the quantum effects cannot be ignored, are obviously excluded
from the current analysis. For systems at sufficently large temperature, the particles
occupy almost exclusively high-energy levels. 
Furthermore, the energy spectrum approaches to be a continuum, as is expected by taking the limit
$\hbar\to0$, and the number of particles per level are very close to zero.

Subsequently, we substitute $f_{nM}\approx0$ into Eq.~(\ref{eq:qHdef}) to obtain:
%
\begin{equation}\label{h-quantic2}
    \mathcal{H}= \sum_{M=1}^{K} \sum_n
    \left[ f_{nM}(t)\ln \left(f_{nM}(t)\right)\right] \delta V_M
    =\sum_{M=1}^{K} \sum_n
    \left[ f_{M}(\epsilon_n,t) \ln \left( f_{M}(\epsilon_n,t)\right)\right] \delta V_M.
\end{equation}
%

Finally, the sum over the quatum energy levels can be replaced by an integral over the velocities
by invoking both the uncertainty principle and the fact that, for free particles, the continuum
energy spectrum
can be written as a function of the velocity. Hence the quantum $H$-functional
transforms, in the classical limit, to:
%
\begin{equation}\label{eq:qHclassLim}
    \mathcal{H}= \sum_{M=1}^{K}\int C'f_M(\vec v,t)\ln\big[C' f_M(\vec v,t)\big] d\vec v,
\end{equation}
%
where $C'$ collects the appropriate constants stemming from writing the energy spectrum as
a function of $\vec v$.

%--------------------------------------

>>>>

\section{Quantum–classical correspondence}\label{sec:qccorrespondence}

In sections \ref{sec:classical} and \ref{sec:quantum}, we saw that the variational
method can be applied to $H$-functionals, which correctly describes the behavior of
classical and quantum dilute gases, with regard to their respective time-evolution.
Both $H$-functionals defined in Eqs.~(\ref{eq:cHdef}) and
(\ref{eq:qHdef}) also recover the well-known distribution functions, either Maxwell-Boltzmann
for a classical gas, Fermi-Dirac for a Fermi gas, or Bose-Einstein for a Bose gas. Nevertheless,
the functionals (\ref{eq:cHdef}) and
(\ref{eq:qHdef}) are seemingly different, and in this section, we will show they are
related by the correspondence principle.

We start by arguing that, in equilibrium, it is straightforward to proof that
Eq.~(\ref{eq:qfneqtwo}) collapses
into Eq.~(\ref{eq:fbardef}) by taking the limit wherein the 
degeneration parameter $\xi\equiv\exp\big(-(\epsilon-\mu)/(k_BT)\big) \gg 1$.
Alternatively, a more general approach to show the quantum–classical correspondence
consists of analyzing the collapse from the quantum to the classical $H$-functionals
within the appropriated limit.
For the case treated here, this limit is $f_{nM} \approx 0$ for several reasons.
Systems at very low temperatures, in which the quantum effects cannot be ignored, are obviously excluded
from the current analysis. In systems at sufficiently high temperatures, the particles
occupy almost exclusively high-energy levels. 
Furthermore, the energy spectrum approaches a continuum, as is expected by taking the limit
$\hbar\to0$, and the number of particles per level is very close to zero.

Subsequently, we substitute $f_{nM}\approx0$ into Eq.~(\ref{eq:qHdef}) to obtain
%
\begin{equation}\label{h-quantic2}
    \mathcal{H}= \sum_{M=1}^{K} \sum_n
    \left[ f_{nM}(t)\ln \left(f_{nM}(t)\right)\right] \delta V_M
    =\sum_{M=1}^{K} \sum_n
    \left[ f_{M}(\epsilon_n,t) \ln \left( f_{M}(\epsilon_n,t)\right)\right] \delta V_M.
\end{equation}
%

Finally, the sum over the quantum energy levels can be replaced by an integral over the velocities
by invoking both the uncertainty principle and the fact that, for free particles, the continuum
energy spectrum
can be written as a function of the velocity. Hence the quantum $H$-functional
transforms, in the classical limit, to
%
\begin{equation}\label{eq:qHclassLim}
    \mathcal{H}= \sum_{M=1}^{K}\int C'f_M(\vec v,t)\ln\big[C' f_M(\vec v,t)\big] d\vec v,
\end{equation}
%
where $C'$ collects the appropriate constants stemming from writing the energy spectrum as
a function of $\vec v$.
